Loop-Aware Optimizations in PyPy's Tracing JIT

### 2012-08-15 talk/dls2012/paper.tex
tune formulations

\begin{abstract}
One of the nice properties of a tracing JIT is that many of its optimizations
are simple requiring one forward pass only. This is not true for loop-invariant code
motion which is a very important optimization for code with tight kernels.
Especially for dynamic languages that typically perform quite a lot of loop invariant
type checking, boxed value unwrapping and virtual method lookups.
In this paper we explain a scheme invented within the context of the LuaJIT project
for making simple optimizations loop-aware by
using a simple pre-processing step on the trace and not changing the
optimizations themselves.
We have implemented the scheme in PyPy's tracing JIT compiler,
where it can give performance improvements of a
factor over two for PyPy's Python JIT executing simple numerical kernels
bringing the performance close to that of compiled C code.
\end{abstract}

@@

We run GCC with -O3 -march=native, disabling the
automatic loop vectorization. In all cases, SSE2 instructions were used for
floating point operations, except Psyco which uses x87 FPU instructions.
% Psyco does not use the x87 FPU: all floating-point arithmetic is done with
% residual calls to C helpers.  These can probably be compiled with SSE2.
% But compiling CPython (and maybe Psyco) for x87 or SSE2 has probably
% no measurable effect.
We also run PyPy with loop peeling optimization and without (but otherwise
identical).

@@

Loop invariant code motion has been part of early compilers in the 1960s and
1970s~\cite{allen_catalogue_1971}. The approach for achieving loop invariant
code motion is typically to perform partial redundancy elimination. The
approach was first proposed by Morel and Renvoise~\cite{morel_global_1979}. It
involves solving data flow problems usually involding bidirection data flow
equations. After improvements~\cite{chow_portable_1984,
dhamdhere_practical_1991} this approach was followed by the work of Knoop
et.al.~\cite{knoop_lazy_1992} who cleanly separated the problem into a backward
and forward data flow analysis. Implementing partial redundancy elimination in
compilers that use SSA form \cite{chow_new_1997} simplified the algorithms
because no iterative data flow analysis is needed any more.

>>>>

\begin{abstract}
One of the nice properties of a tracing JIT is that many of its optimizations
are simple requiring one forward pass only. This is not true for loop-invariant code
motion which is a very important optimization for code with tight kernels.
Especially for dynamic languages that typically perform quite a lot of loop invariant
type checking, boxed value unwrapping and virtual method lookups.
In this paper we explain a scheme pioneered within the context of the LuaJIT project
for making simple optimizations loop-aware by
using a simple pre-processing step on the trace and not changing the
optimizations themselves.
We have implemented the scheme in PyPy's tracing JIT compiler,
where it can give performance improvements of a
factor over two for PyPy's Python JIT executing simple numerical kernels
bringing the performance close to that of compiled C code.
\end{abstract}

@@

We run GCC with -O3 -march=native, disabling the
automatic loop vectorization. In all cases, SSE2 instructions were used for
floating point operations, except Psyco which uses x87 FPU instructions.
We also run PyPy with loop peeling optimization and without (but otherwise
identical).

@@

Loop invariant code motion has been part of early compilers in the 1960s and
1970s~\cite{allen_catalogue_1971}. A common approach for achieving loop invariant
code motion is to perform partial redundancy elimination. The
approach was first proposed by Morel and Renvoise~\cite{morel_global_1979}. It
involves solving data flow problems of bidirectional data flow
equations. After improvements~\cite{chow_portable_1984,
dhamdhere_practical_1991} this approach was followed by the work of Knoop
et.al.~\cite{knoop_lazy_1992} who cleanly separated the problem into a backward
and forward data flow analysis. Implementing partial redundancy elimination in
compilers that use SSA form \cite{chow_new_1997} simplified the algorithms
because no iterative data flow analysis is needed any more.
